A comps-based approach for interpreting tree-based predictions with an application to the NFL draft

### 2025-06-24 articles/arxiv/paper.tex
add paragraph explaining adjusted QBR

@@

To train the random forest model, we first implemented a random 70\%/30\% train/test split. We chose $B$ = 500 decision trees, each trained on a bootstramp sample of the data, with replacement. To determine the number of variables to consider at each node and the maximum tree size, we used the R package tuneRanger \citep{probst_tuneranger_2018}, which uses model-based optimization of out-of-bag prediction error to tune the hyperparameters. After 100 iterations of the tuning algorithm, we estimated the optimal number of variables to consider at each node to be X and the optimal minimum node size to be XX.

@@

Figure \ref{fig:importance} shows the importance of each feature learned by the random forest model. The model's test RMSE was 9.89 units of QBR, corresponding to explaining 35.1\% of the variance of QBR in the test set. Average yards per season, completions per season and touchdowns per season in college stand out as the most important predictors of NFL success. Note that these three variables are highly correlated with each other---each pairwise correlation is at least 0.94. The next most important variables are final-season Heisman voting, average passing attempts per season and final-season passer rating.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.75\linewidth]{../../figures/variable_importance.pdf}
    \caption{\textit{Variable importance plot for random forest trained to predict NFL QBR for college quarterbacks.}}
    \label{fig:importance}
\end{figure}

Figure \ref{fig:3d-plot} visualizes the prediction function learned by the random forest by plotting predicted QBR against career passing yards per season and final-season passer rating. These are two of the most important predictors in the model, and they have a relatively low correlation. The extreme outlier by passing yards per season is Bailey Zappe, who played only one FBS season at Western Kentucky in Conference USA and was selected by the New England Patriots in the fourth round of the 2022 NFL Draft. The predictions of the random forest model are conservative, with extremely few quarterbacks predicted to exceed a QBR of 40. This speaks to the large number of college quarterbacks who achieve zero QBR in the NFL and the difficulty of distinguishing between prospects based on college performance statistics alone.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.75\linewidth]{../../figures/3d_plot.pdf}
    \caption{\textit{Predicted QBR plotted against average passing yards per season and final-season passer rating.}}
    \label{fig:3d-plot}
\end{figure}

Figure \ref{fig:predicted-vs-actuals} shows actual vs. predicted QBR for all quarterbacks in the dataset. Recall that 91\% of actual QBRs are zero, which the visualization does not make clear. The errors are very right-skewed, and this illustrates why no college quarterback exceed a QBR prediction of 50.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\linewidth]{../../figures/predicted_vs_actuals.pdf}
  \caption{\textit{Predicted (from random forest) and actual QBR for all quarterbacks in the dataset. The vats majority (91\%) of actual QBRs are zero, meaning that the points along the x-axis represent 91\% of the data.}}
  \label{fig:predicted-vs-actuals}
\end{figure}

@@

Table \ref{tab:top-ten} shows the QBR predictions for the top 10 predicted quarterbacks in the 2025 draft. Cam Ward is predicted to have the highest QBR, which matches the common consensus that he was the best quarterback in this draft class and the fact he was taken first overall in the draft. Dillon Gabriel and Shedeur Sanders were projected to have the second and third highest QBRs, respectively. While Dillon Gabriel was not previously expected to be drafted as high as Shedeur Sanders, he had a very consistent college career and finished the highest out of any quarterback in Heisman voting in 2025. Jackson Dart's draft stock increased substantially after the Scouting Combine and his pro day.

\begin{table}[H]
  \centering
  \begin{tabular}{l|r}
    Player & Predicted QBR\\
    \hline
    \input{../../tables/top_ten.tex}
  \end{tabular}
  \caption{\textit{Top ten career-average QBR predictions for quarterback prospects in the 2025 draft class. This includes only draft-eligible players who played at an NCAA Division I FBS school during the 2024 season.}}
  \label{tab:top-ten}
\end{table}

@@

It is worth noting that both Ward and Sanders started their college careers at teams that are not in the NCAA Division I FBS, as Ward played his first season at Incarnate Word and Sanders played his first two seasons at Jackson State. As a result, their college statistics used in the model do not include their statistics from these schools. We observe that Ward has the highest QBR prediction, which intuitively makes sense because he is widely regarded as the best quarterback in this year's draft class. Gabriel has a slightly higher QBR prediction than Sanders although Sanders was considered a better prospect. However, Gabriel finished higher than Sanders in Heisman voting, and Oregon had a much better record than Colorado, and Gabriel was ultimately taken before Sanders in the draft. Dart's relatively low prediction compared to the other three makes sense because he wasn't considered a top quarterback prospect until his performance in the NFL Combine, which this model does not take into consideration. He also performed well in his interviews with teams, which is not a quantifiable measure.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.24\linewidth]{../../figures/prospect_histogram_ward.pdf}
    \includegraphics[width=0.24\linewidth]{../../figures/prospect_histogram_sanders.pdf}
    \includegraphics[width=0.24\linewidth]{../../figures/prospect_histogram_gabriel.pdf}
    \includegraphics[width=0.24\linewidth]{../../figures/prospect_histogram_dart.pdf} \\
    \includegraphics[width=0.24\linewidth]{../../figures/prospect_similarity_ward.pdf}
    \includegraphics[width=0.24\linewidth]{../../figures/prospect_similarity_sanders.pdf}
    \includegraphics[width=0.24\linewidth]{../../figures/prospect_similarity_gabriel.pdf}
    \includegraphics[width=0.24\linewidth]{../../figures/prospect_similarity_dart.pdf}
    \caption{\textit{Top: Histogram of training players' QBR values weighted by similarity to the reference prospect. The vertical green line annotates each prospect's predicted QBR, which matches the mean of the histograph distribution. Bottom: Scatter plot showing individual similarity scores and QBR for historical prospects.}}
    \label{fig:prospect-plots}
\end{figure}

@@

In the application to the 2025 NFL draft, the random forest's predictions explain close to one third of the out-of-sample variance in QBR, which highlights some limitations of statistical forecasting when scouting draft prospects. The variable importance rankings show the college statistics that have a tendency to translate into better NFL performance. However, these two things also highlight the importance of non-quantifiable factors---such as attitude, coachability, off-the-field issues, adaptability, and leadership.

>>>>

Since it is possible for a quarterback with very few pass attempts to have a high QBR, we adjusted the average QBR by regressing it toward zero based on attempts per season. We used attempts per season rather than career attempts to prevent discrimination against players who have not yet played many seasons. We used the formula $QBR_{adj}=/frac{avg_att*\bar{QBR}}{avg_att+c}$ for regression, where c is a constant. We used a nonlinear least squares model with a formula of $QBR2~/frac{(Att * QBR)}{(Att + c)}$, where Att is that season's total attempts, QBR is that season's QBR, and QBR2 is the next season's QBR, to find c, resulting in a value of 22.50.

@@

To train the random forest model, we first implemented a random 70\%/30\% train/test split. We chose $B$ = 500 decision trees, each trained on a bootstramp sample of the data, with replacement. To determine the number of variables to consider at each node and the maximum tree size, we used the R package tuneRanger \citep{probst_tuneranger_2018}, which uses model-based optimization of out-of-bag prediction error to tune the hyperparameters. After 100 iterations of the tuning algorithm, we estimated the optimal number of variables to consider at each node to be 3 and the optimal minimum node size to be 21.

@@

Figure \ref{fig:importance} shows the importance of each feature learned by the random forest model. The model's test RMSE was 8.61 units of QBR, corresponding to explaining 33.3\% of the variance of QBR in the test set. Average yards per season, completions per season and touchdowns per season in college stand out as the most important predictors of NFL success. Note that these three variables are highly correlated with each other---each pairwise correlation is at least 0.94. The next most important variables are final-season Heisman voting, average passing attempts per season and final-season passer rating.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.75\linewidth]{../../figures/variable_importance.pdf}
    \caption{\textit{Variable importance plot for random forest trained to predict adjusted NFL QBR for college quarterbacks.}}
    \label{fig:importance}
\end{figure}

Figure \ref{fig:3d-plot} visualizes the prediction function learned by the random forest by plotting predicted adjusted QBR against career passing yards per season and final-season passer rating. These are two of the most important predictors in the model, and they have a relatively low correlation. The extreme outlier by passing yards per season is Bailey Zappe, who played only one FBS season at Western Kentucky in Conference USA and was selected by the New England Patriots in the fourth round of the 2022 NFL Draft. The predictions of the random forest model are conservative, with extremely few quarterbacks predicted to exceed an adjusted QBR of 40. This speaks to the large number of college quarterbacks who achieve zero QBR in the NFL and the difficulty of distinguishing between prospects based on college performance statistics alone.

\begin{figure}[H]
    \centering
    \includegraphics[width=0.75\linewidth]{../../figures/3d_plot.pdf}
    \caption{\textit{Predicted adjusted QBR plotted against average passing yards per season and final-season passer rating.}}
    \label{fig:3d-plot}
\end{figure}

Figure \ref{fig:predicted-vs-actuals} shows actual vs. predicted adjusted QBR for all quarterbacks in the dataset. Recall that 91\% of actual QBRs are zero, which the visualization does not make clear. The errors are very right-skewed, and this illustrates why no college quarterback exceed an adjusted QBR prediction of 50.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.5\linewidth]{../../figures/predicted_vs_actuals.pdf}
  \caption{\textit{Predicted (from random forest) and actual adjusted QBR for all quarterbacks in the dataset. The vats majority (91\%) of actual QBRs are zero, meaning that the points along the x-axis represent 91\% of the data.}}
  \label{fig:predicted-vs-actuals}
\end{figure}

@@

Table \ref{tab:top-ten} shows the adjusted QBR predictions for the top 10 predicted quarterbacks in the 2025 draft. Cam Ward is predicted to have the highest adjusted QBR, which matches the common consensus that he was the best quarterback in this draft class and the fact he was taken first overall in the draft. Dillon Gabriel and Shedeur Sanders were projected to have the second and third highest adjusted QBRs, respectively. While Dillon Gabriel was not previously expected to be drafted as high as Shedeur Sanders, he had a very consistent college career and finished the highest out of any quarterback in Heisman voting in 2025. Jackson Dart's draft stock increased substantially after the Scouting Combine and his pro day.

\begin{table}[H]
  \centering
  \begin{tabular}{l|r}
    Player & Predicted QBR\\
    \hline
    \input{../../tables/top_ten.tex}
  \end{tabular}
  \caption{\textit{Top ten career-average adjusted QBR predictions for quarterback prospects in the 2025 draft class. This includes only draft-eligible players who played at an NCAA Division I FBS school during the 2024 season.}}
  \label{tab:top-ten}
\end{table}

@@

It is worth noting that both Ward and Sanders started their college careers at teams that are not in the NCAA Division I FBS, as Ward played his first season at Incarnate Word and Sanders played his first two seasons at Jackson State. As a result, their college statistics used in the model do not include their statistics from these schools. We observe that Ward has the highest adjusted QBR prediction, which intuitively makes sense because he is widely regarded as the best quarterback in this year's draft class. Gabriel has a slightly higher QBR prediction than Sanders although Sanders was considered a better prospect. However, Gabriel finished higher than Sanders in Heisman voting, and Oregon had a much better record than Colorado, and Gabriel was ultimately taken before Sanders in the draft. Dart's relatively low prediction compared to the other three makes sense because he wasn't considered a top quarterback prospect until his performance in the NFL Combine, which this model does not take into consideration. He also performed well in his interviews with teams, which is not a quantifiable measure.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.24\linewidth]{../../figures/prospect_histogram_ward.pdf}
    \includegraphics[width=0.24\linewidth]{../../figures/prospect_histogram_sanders.pdf}
    \includegraphics[width=0.24\linewidth]{../../figures/prospect_histogram_gabriel.pdf}
    \includegraphics[width=0.24\linewidth]{../../figures/prospect_histogram_dart.pdf} \\
    \includegraphics[width=0.24\linewidth]{../../figures/prospect_similarity_ward.pdf}
    \includegraphics[width=0.24\linewidth]{../../figures/prospect_similarity_sanders.pdf}
    \includegraphics[width=0.24\linewidth]{../../figures/prospect_similarity_gabriel.pdf}
    \includegraphics[width=0.24\linewidth]{../../figures/prospect_similarity_dart.pdf}
    \caption{\textit{Top: Histogram of training players' adjusted QBR values weighted by similarity to the reference prospect. The vertical green line annotates each prospect's predicted adjusted QBR, which matches the mean of the histograph distribution. Bottom: Scatter plot showing individual similarity scores and adjusted QBR for historical prospects.}}
    \label{fig:prospect-plots}
\end{figure}

@@

In the application to the 2025 NFL draft, the random forest's predictions explain close to one third of the out-of-sample variance in adjusted QBR, which highlights some limitations of statistical forecasting when scouting draft prospects. The variable importance rankings show the college statistics that have a tendency to translate into better NFL performance. However, these two things also highlight the importance of non-quantifiable factors---such as attitude, coachability, off-the-field issues, adaptability, and leadership.
